Global patterns of forest autotrophic carbon fluxes

### 2020-05-06 manuscript/manuscript draft.Rmd
worked on discussion, bringing in more refs
#69 , #43

Length of the growing season was estimated to the nearest month, where growing season months were defined as months with mean minimum temperature > 0.5$^\circ$C. We experimented with a definion of growing season months including a moisture index, defined as (MAT - PET)/PET, > -0.95 (Kerkhoff et al. 2005; see also Michaletz et al. 2014-**put in these refs**). However, we found that including a moisture index had **no** effect on the estimates of growing season length, and so chose to exclude it.  (**Becky, was it really no effect? or minimal?**) Monthly data for PET, precipitation, and temperature from the CRU dataset v 4.03 [@harris_updated_2014], and solar radiation from WorldClim2 [@fick_worldclim_2017] were used to calculate mean monthly PET, precipitation, temperature and solar radiation during the growing season. Total growing season precipitation and solar radiation were also calculated.

@@

**This paragraph needs work, but I'm not currently sure how to fix it. It reads more as results than discussion, and doesn't say anything really profound. Depending on what happens with [issue #70](https://github.com/forc-db/Global_Productivity/issues/70), we might somehow rework/rearrange this paragraph and the previous two. **
The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, and secondarily by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means, it is not appropriate to attempt to identify any one mean annual climate variable as a mechanistic driver of FACF. However, it remains informative to consider these relationships. We find that temperature-related climate variables ($MAT$, temperature seasonality, ...**LIST**) explain the highest proportion of variability in FACF, and among these, $MAT$ is generally the best predictor--perhaps because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables are extracted from global gridded data products (Table S1). The effects of temperature are modified by moisture availability, with reduced FACF under hot and dry conditions (*i.e.*, high PET, high deficit; Fig. 4c-d) and sometimes under very high precipitaiton (Figs. 3, 4b). Negative effects of very high precipitation on FACF have been observed previously (**REFS**) and are attributable to nutrient and light limitations (**REFS**). Thus, although temperature and water availability jointly and interactively drive global-scale patterns of FACF....

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and is minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions (Table 1, *H5*; **REFS**). We found that the sort of simple metric needed to define growing season at a global scale [@kerkhoff_plant_2005;@michaletz_convergence_2014] was rough with respect to temperature and problematic with respect to moisture because it doesn't capture temporal lags between precipitation and plant water use caused by storage in soil or snow. (**probably better to get into this in the methoods**). A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosyntheiss or respiration by high-latitude forests (*[ Enquist et al. 2007 GCB- but check]*). However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains very crude relative to the the timescales on which climate affects plant metabolism and does not advance statistical predictive power. Rather, mechanistic accounting for climatic effects on global FACF patterns requires models representing physiologically meaningful timescales (*e.g.*, **refs**).

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ [@li_mapping_2019] and $ANPP_{woody stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates are not universal [@rutishauser_testing_2020] (**MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

## misc content for discussion
- the observed positive interaction between MAT and MAP for $ANPP_woody$ is consitent with the @taylor_temperature_2017 analysis showing such an interaction for $ANPP$ in tropical forests. Similar to their analysis, we find a cross-over point at ~20C. However, we don't find such an interaction for $ANPP$, and we show a contrasting negative interaction for $NPP$. Some of this is may be stochasitic/ driven by composition of the dataset, and the interactions we observe are not internally consistent.

-results are consistent with *Muller-Landau et al., in review*

>>>>

Length of the growing season was estimated to the nearest month, where growing season months were defined as months with mean minimum temperature > 0.5$^\circ$C. We experimented with a definion of growing season months including a moisture index, defined as (MAT - PET)/PET, > -0.95 [@kerkhoff_plant_2005; see also @michaletz_convergence_2014]. However, we found that including a moisture index had **no** effect on the estimates of growing season length, and so chose to exclude it.  (**Becky, was it really no effect? or minimal?**) Monthly data for PET, precipitation, and temperature from the CRU dataset v 4.03 [@harris_updated_2014], and solar radiation from WorldClim2 [@fick_worldclim_2017] were used to calculate mean monthly PET, precipitation, temperature and solar radiation during the growing season. Total growing season precipitation and solar radiation were also calculated.

@@

The latitudinal gradient in FACF (Fig. 2) is driven primarily by temperature-related climate variables, the effects of which are moderated by moisture availability (Table 1, *H2-H3*; Figs. 3-4). Indeed, MAT and MAP have long been identified as primary global-scale drivers of FACF [*Lieth 1973: doi 10.1007/BF01536729*; **REFS**; @taylor_temperature_2017]. It is not appropriate to attempt to identify individual mean annual climate variables as mechanistic drivers of FACF because many climate variables co-vary across the latitudinal gradient (Fig. S2), because climatic drivers affect forest carbon flux on much shorter time scales than can be captured by annual climate summary variables, and because both climatic conditions and C flux vary intra- and inter-annually around the long-term means. However, it remains informative to consider these relationships. Among the  temperature-related climate variables ($MAT$, temperature seasonality, annual temperature range, annual frost days, length of growing season, and $PET$; Table S1), $MAT$ is generally the most strongly correlated with FACF (Table S2)--perhaps in part because site-specific MAT is recorded for the majority of sites in ForC, whereas other variables were extracted from global gridded data products. This finding supports the continued focus on MAT as a primary--*albiet* not mechanistic--correlate of FACF. The effects of $MAT$ are modified by moisture availability, with reduced FACF under relatively dry conditions (*i.e.*, low precipitation; high vapour pressure deficit) and sometimes under very high precipitation (Figs. 3-4). The observed positive interaction between MAT and MAP for $ANPP_{woody.stem}$ on the global scale (Fig. 3) is consistent with an analysis showing a similar interaction for $ANPP$ in tropical forests, also with a cross-over point at ~20$\circ$C [@taylor_temperature_2017]. However, we detect no such interaction for $ANPP$ or most other FACF, and we find a contrasting negative interaction for $NPP$ (Fig. 3), suggesting that more data are required to sort out potential differences in the interactive effects of MAT and MAP on FACF in the tropics.

FACF are negatively correlated with temperature seasonality (Table 1, *H4*; Fig. 4e), and are minimal during cold- or dry- dormant seasons. To account for this, a number of analyses seeking to characterize global-scale effects of climate on productivity have examined the relationship of C flux per month of the growing season with growing season climatic conditions [Table 1, *H5*; @kerkhoff_plant_2005; *Anderson et al. 2006: doi:10.1111/j.1461-0248.2006.00914.x *; *Enquist et al. 2007: doi: 10.1111/j.1365-2486.2006.01222.x*;  @michaletz_convergence_2014]. We found that the sort of simple metric needed to define growing season at a global scale [@kerkhoff_plant_2005] was rough with respect to temperature and problematic with respect to moisture because it doesn't capture temporal lags between precipitation and plant water use caused by storage in soil or snow. A temperature-defined growing season length had stong positive correlation with FACF (Fig. 4f), but explained less variation than $MAT$. Dividing FACFs by growing season length to yield FACF per growing season month removed the majority of climate-related variation, supporting the idea that the latitudinal gradient in FACF is attributable more to shorter growing seasons at high latitudes than to inherently lower rates of photosyntheiss or respiration by high-latitude forests (*Enquist et al. 2007: doi: 10.1111/j.1365-2486.2006.01222.x*). However, there remained a number of significant correlations with growing season climatic conditions, suggesting that climatic conditions remain influential within the growing season. We conclude that while correcting for growing season length takes analyses a step closer to mechanistic linkage of instantaneous C flux rates to environmental conditions, it remains very crude relative to the the timescales on which climate affects plant metabolism and does not advance statistical predictive power. Rather, mechanistic accounting for climatic effects on global FACF patterns requires models representing physiologically meaningful timescales (*e.g.*, **refs**).

Our analysis clarifies how FACF vary with latitude and climate on a global scale, with some important implications for how forest carbon cycling relates to climate and, by extension, how it is likely to respond to climatic warming. Our findings show that higher temperatures with similar moisture availability result in a generalized acceleration of FACF (Figs. 2-3). This is consistent with observations of continental- to global-scale increases in $GPP$ [@li_mapping_2019] and $ANPP_{woody stem}$ [@brienen_long-term_2015; @hubau_asynchronous_2020], along with some C cycle components not considered here--tree mortality [@brienen_long-term_2015; @mcdowell_drivers_2018], soil respiration [@bond-lamberty_global_2010], and heterotrophic soil respiration [@bond-lamberty_globally_2018]. However, increasing C flux rates by no means universal [@rutishauser_testing_2020] (**MORE REFS**). This is likely because factors other than rising temperatures are at play, including changes to other aspects of climate, atmospheric pollution (CO~2~, SO~2~, NO~x~), and local disturbances. Morevoer, forest ecosystem responses to climatic changes outside the temperature range to which forest communities are adapated and acclimatized will not necessarily parallel responses across across geographic gradients in climate. Nevertheless, as we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.
